Fluent statistical computing interfaces for biological data analysis

### 2020-07-26 Rmd/02-plyranges.Rmd
typos

We have created a genomic DSL called **plyranges** that reformulates
notions from existing genomic algebras and embeds them in R as a
genomic extension of **dplyr**. By analogy, **plyranges** is to the
genomic algebra, as **dply**` is to the relational algebra.  The
**plyranges`**Bioconductor package implements the language on top of a
key subset of Bioconductor data structures and thus fully integrates
with the Bioconductor framework, gaining access to its scalable data
representations and sophisticated statistical methods.

@@

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
   Aggregate & \textt{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \textt{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
   \hline
   &  \textbf{\textt{mutate()}} & modifies any column \\
   & \textbf{\textt{select()}} & select columns \\
  Modify (Unary) & \textbf{\textt{arrange()}} & sort by columns \\
   & \textt{stretch()} & extend range by fixed amount \\
   &  \textt{shift\_(direction)} & shift coordinates \\
   & \textt{flank\_(direction)} & generate flanking regions \\
   & \textt{\%intersection\% } & row-wise intersection \\
   & \textt{\%union\%} & row-wise union \\
   & \textt{compute\_coverage} & coverage over all ranges \\
  Modify (Binary) &  \textt{\%setdiff\%} & row-wise set difference \\
   & \textt{between()} & row-wise gap range \\
   & \textt{span()} & row-wise spanning range \\
   \hline
    & \textt{join\_overlap\_*()} & merge by overlapping ranges \\
    & \textt{join\_nearest} & merge by nearest neighbor ranges \\
    & \textt{join\_follow} & merge by following ranges \\
    Merge & \textt{join\_precedes} & merge by preceding ranges \\
    & \textt{union\_ranges} & range-wise union \\
    & \textt{intersect\_ranges} & range-wise intersect \\
    & \textt{setdiff\_ranges} & range-wise set difference \\
    & \textt{complement\_ranges} & range-wise set complement \\
  \hline
   & \textt{anchor\_direction()} & fix coordinates at direction \\
  Operate & \textbf{\textt{group\_by()}} & partition by column(s)  \\ 
   & \textt{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\textt{filter()}} & subset rows \\
  Restrict & \textt{filter\_by\_overlaps()} & subset by overlap \\
    & \textt{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \textbf{plyranges} grammar. The core verbs are
briefly described and categorized into one of the following higher level 
categories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
its origin is from the \textbf{dplyr} grammar.}\label{tab:grammar}
\end{table}

@@

We aim to continue developing the **plyranges** package and to extend it
for use with more complex data structures, such as the
_SummarizedExperiment_ class, the core Bioconductor data structure for
representing experimental results (e.g., counts) from multiple sample
experiments in conjunction with feature and sample metadata. Although,
the _SummarizedExperiment_ is not strictly tidy, it does consist of 
three tidy data structures that are related by feature and sample identifiers. 
Therefore, the grammar and design of the **plryanges** DSL is naturally 
extensible to the _SummarizedExperiment_.

>>>>

We have created a genomic DSL called **plyranges** that reformulates
notions from existing genomic algebras and embeds them in R as a
genomic extension of **dplyr**. By analogy, **plyranges** is to the
genomic algebra, as **dplyr**` is to the relational algebra.  The
**plyranges`**Bioconductor package implements the language on top of a
key subset of Bioconductor data structures and thus fully integrates
with the Bioconductor framework, gaining access to its scalable data
representations and sophisticated statistical methods.

@@

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\texttt{summarize()}} & aggregate over column(s) \\ 
   Aggregate & \texttt{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \texttt{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
   \hline
   &  \textbf{\texttt{mutate()}} & modifies any column \\
   & \textbf{\texttt{select()}} & select columns \\
  Modify (Unary) & \textbf{\texttt{arrange()}} & sort by columns \\
   & \texttt{stretch()} & extend range by fixed amount \\
   &  \texttt{shift\_(direction)} & shift coordinates \\
   & \texttt{flank\_(direction)} & generate flanking regions \\
   & \texttt{\%intersection\% } & row-wise intersection \\
   & \texttt{\%union\%} & row-wise union \\
   & \texttt{compute\_coverage} & coverage over all ranges \\
  Modify (Binary) &  \texttt{\%setdiff\%} & row-wise set difference \\
   & \texttt{between()} & row-wise gap range \\
   & \texttt{span()} & row-wise spanning range \\
   \hline
    & \texttt{join\_overlap\_*()} & merge by overlapping ranges \\
    & \texttt{join\_nearest} & merge by nearest neighbor ranges \\
    & \texttt{join\_follow} & merge by following ranges \\
    Merge & \texttt{join\_precedes} & merge by preceding ranges \\
    & \texttt{union\_ranges} & range-wise union \\
    & \texttt{intersect\_ranges} & range-wise intersect \\
    & \texttt{setdiff\_ranges} & range-wise set difference \\
    & \texttt{complement\_ranges} & range-wise set complement \\
  \hline
   & \texttt{anchor\_direction()} & fix coordinates at direction \\
  Operate & \textbf{\texttt{group\_by()}} & partition by column(s)  \\ 
   & \texttt{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\texttt{filter()}} & subset rows \\
  Restrict & \texttt{filter\_by\_overlaps()} & subset by overlap \\
    & \texttt{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \textbf{plyranges} grammar. The core verbs are
briefly described and categorized into one of the following higher level 
categories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
its origin is from the \textbf{dplyr} grammar.}\label{tab:grammar}
\end{table}

@@

We aim to continue developing the **plyranges** package and to extend it
for use with more complex data structures, such as the
_SummarizedExperiment_ class, the core Bioconductor data structure for
representing experimental results (e.g., counts) from multiple sample
experiments in conjunction with feature and sample metadata. Although,
the _SummarizedExperiment_ is not strictly tidy, it does consist of 
three tidy data structures that are related by feature and sample identifiers. 
Therefore, the grammar and design of the **plyranges** DSL is naturally 
extensible to the _SummarizedExperiment_.

### 2020-07-26 Rmd/03-fluentGenomics.Rmd
typos

We have already used **plyrange**  a number of times above, to `filter()`,
`mutate()`, and `select()` on _GRanges_ objects, as well as ensuring the correct
genome annotation and style has been used. The **plyranges** package provides a
grammar for performing transformations of genomic data [@Lee2019]. Computations
resulting from compositions of **plyranges** "verbs" are performed using
underlying, highly optimized range operations in the **GenomicRanges** package
[@granges].

>>>>

We have already used **plyranges**  a number of times above, to `filter()`,
`mutate()`, and `select()` on _GRanges_ objects, as well as ensuring the correct
genome annotation and style has been used. The **plyranges** package provides a
grammar for performing transformations of genomic data [@Lee2019]. Computations
resulting from compositions of **plyranges** "verbs" are performed using
underlying, highly optimized range operations in the **GenomicRanges** package
[@granges].

### 2020-07-26 Rmd/05-embeddings.Rmd
typos

These checks are done visually using our side-by-side tour and embedding views.
In the simplest case, a local continuity check can be assessed via one to
one linked brushing from the embedding to the tour view. Similarly, diffusions
are identified from linked brushing on the tour view, higlighting points
in the embedding view. Section <> details

@@

In future implementations it will be possible to shift from one to one linked
brushing to one to many linked brushing. This form of brushing acts directly on
a $k$-nearest neighbours ($k$-nn) graph computed from a
reference dataset: when a user brushes over a region in the embedding,
all the points that match the graphs edges are selected on the 
corresponding tour view.  The reference data set for computing
nearest neighbours (for example a distance matrix, or the complete data matrix)
can be independent of the tour or embedding views. The estimation of neighbours
can be computed effeciently with the `FNN` or `RcppAnnoy` packages
on the server side [@fnn-pkg; @rcpp-annoy-pkg]. In place of highlighting, one 
could use opacity or binned colour scales to encode distances or ranks instead 
of the neighbouring points. Brush composition, such as 'and', 'or',
or 'not' brushes, could be used to interrogate mismatches between the 
$k$-nn graphs estimated from both the embedding and tour views.

>>>>

These checks are done visually using our side-by-side tour and embedding views.
In the simplest case, a local continuity check can be assessed via one to
one linked brushing from the embedding to the tour view. Similarly, diffusions
are identified from linked brushing on the tour view, highlighting points
in the embedding view. Section <> details

@@

In future implementations it will be possible to shift from one to one linked
brushing to one to many linked brushing. This form of brushing acts directly on
a $k$-nearest neighbours ($k$-nn) graph computed from a
reference dataset: when a user brushes over a region in the embedding,
all the points that match the graphs edges are selected on the 
corresponding tour view.  The reference data set for computing
nearest neighbours (for example a distance matrix, or the complete data matrix)
can be independent of the tour or embedding views. The estimation of neighbours
can be computed efficiently with the `FNN` or `RcppAnnoy` packages
on the server side [@fnn-pkg; @rcpp-annoy-pkg]. In place of highlighting, one 
could use opacity or binned colour scales to encode distances or ranks instead 
of the neighbouring points. Brush composition, such as 'and', 'or',
or 'not' brushes, could be used to interrogate mismatches between the 
$k$-nn graphs estimated from both the embedding and tour views.
